Evaluation of +S2 sea ice forecast and some other better title

### 2025-10-06 analysis/paper/access-ice.qmd
Formatting

```{r mirai_daemons}
if (!mirai::daemons_set()) invisible(daemons(6))
```

@@

```{r range_max_lag}
range_max_lag <- hindcast_extent[mday(time) == 2, .SD[which.max(lag)], 
                                 by = .(forecast_time, model)] |> 
  _[, .(max = max(lag),
        min = min(lag)),
    by = model]
```

@@

hindcast_extent[, .(model, forecast_time, member, time, Forecast = aice)] |> 
  extent_daily[i = _, on = c("time")] |> 
  _[, Difference := Forecast - aice] |> 
  _[, aice := NULL] |> 
  tidyfast::dt_pivot_longer(cols = c(Forecast, Difference)) |> 
  _[, average(value), 
    by = .(dataset, 
           model, 
           measure = name,
           forecast_time = update(forecast_time, year = 2001),
           lag = as.numeric(as.Date(time) - forecast_time))] |> 
  _[, time := forecast_time + lag] |>
  _[, time2 := update(time, year = 2001)] |>
  _[, month := lubridate::month(forecast_time, label = TRUE)] |> 
  _[, model := factor(model, levels = c("S1", "S2"))] |> 
  _[, measure := factor(measure, levels = c("Forecast", "Difference"))] |>

  ggplot(aes(time2, estimate)) +
  geom_line(data = extent_climatology[dataset == which_dataset] |> 
              _[, measure := factor("Forecast", levels = c("Forecast", "Difference"))], 
            aes(x = as.Date(time), colour = dataset, 
                group = dataset)) +

  geom_hline(data = data.table(y = 0, measure = factor("Difference", levels = c("Forecast", "Difference"))), aes(yintercept = 0)) +
  geom_line(aes(group = interaction(forecast_time, time != time2), colour = model)) +
  geom_point(data = \(x) x[mday(time) == 2, .SD[which.max(lag)], 
                           by = .(forecast_time, model, measure)], 
             aes(fill = model), 
             shape = 24, size = 2.2) +
  geom_point(data = \(x) x[lag == 1], 
             aes(fill = model), 
             shape = 21, size = 2) +

  scale_x_date(NULL, date_breaks = "1 month", 
               date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = labels_extent) +
  scale_color_models + 
  scale_fill_models +
  guides(colour = "none", fill = "none") +
  # scale_color_manual(values = cetcolor::cet_pal(12, name = "c2"), 
  #                    aesthetics = c("fill", "color"), guide = "none") +
  facet_grid(measure ~ model, scales = "free_y", labeller = labeller(model = labels_models)) +
  tag_facets(tag = "rc") +
  coord_cartesian(clip = "off")
```

@@

```{r extent-delta-compute}
n <- 366*3
w <- 11

dif_hindcast <- hindcast_extent |> 
  copy() |> 
  _[order(time)] |> 
  _[, .(aice = mean(aice)), by = .(model, forecast_time, time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), 
    by = .(model, forecast_time)] |> 
  _[, dif := frollmean(dif, n = w, align = "center"), by = .(model, forecast_time)] |> 
  na.omit() |> 
  _[, average(dif), by = .(model, month(forecast_time), time = update(time, year = 2000))]

dif_extent <- extent_daily |> 
  _[order(time)] |> 
  _[, dif := c(NA, diff(aice)/diff(as.numeric(as.Date(time)))), by = .(dataset)] |> 
  na.omit() |> 
  _[, dif := frollmean(dif, n = w, align = "center"), by = .(dataset)] |> 
  _[, average(dif), by = .(dataset, time = update(time, year = 2000))] |> 
  _[dataset == which_dataset]

@@

dif_extent |> 
  copy() |> 
  ggplot(aes(time, estimate)) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.2) +
  geom_line(aes(color = dataset)) +
  geom_line(data = dif_hindcast,
            aes(color = model, group = interaction(model, month, month > month(time)))) +
  scale_x_datetime(NULL, date_breaks = "1 month", 
                   date_labels = "%b", expand = c(0, 0)) +
  scale_y_continuous(NULL, labels = \(x) labels_extent(x, units = "M km²/day")) +
  scale_color_models +
  coord_cartesian(xlim = as.POSIXct(c("2000-01-01", "2000-12-31"), tz = "UTC"))
```

```{r hindcast_clim}
zeropad <- function(x, n = 2) formatC(x, width = 2, flag = "0")
monnb <- function(d) { lt <- as.POSIXlt(as.Date(d, origin="1900-01-01")); 
lt$year*12 + lt$mon } 
mondf <- function(d1, d2) { monnb(d2) - monnb(d1) }

@@

hindcast_clim <- lapply(setNames(months, months), \(month) {
  lapply(c("S1", "S2"), \(model) {
    here::here(glue::glue("data/derived/climatology/{model}/{zeropad(month)}/em.nc")) |> 
      cdo_selmonth(month) |> 
      cdo_execute() |> 
      ReadNetCDF("aice")  |> 
      _[, model := model]
  }) |> 
    rbindlist()
}) |> 
  rbindlist(idcol = "forecast_month") |> 
  _[, forecast_month := as.numeric(forecast_month)]

hindcast_clim <- hindcast_clim |> 
  copy() |> 
  _[, .(aice = mean(aice)), by = .(month(time), ygrid, xgrid, forecast_month, model)] |> 
  _[forecast_month == month]

@@

s <- lapply(c("S1", "S2"), \(x) {
  g <- hindcast_clim |>
    _[model == x] |>
    # dcast(time + xgrid + ygrid ~ model, value.var = "aice") |> 
    # setnames(which_dataset, "obs") |>
    ggplot(aes(xgrid, ygrid)) +
    geom_contour_fill(aes(z = aice - obs, fill = after_stat(level)), 
                      breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
    scale_fill_divergent_discretised(paste0("Bias with respect to ",
                                            labels_models[which_dataset]),
                                     low = trans_pink,
                                     high = trans_blue) +
    # geom_polygon(data = ice_shelves_nsicd, aes(X, Y, group = group),
    #              colour = "black", fill = "#FAFAFA") +
    geomcoord_antarctica +
    geom_antarctica_fill +
    facet_wrap(~ month + model , labeller = labeller(month = labels_month)) +
    wide_legend +
    tag_facets()

  print(g)

})

@@

monthly_extent <- extent_daily |> 
  _[dataset == which_dataset] |> 
  _[aice == 0, aice := NA]  |> 
  _[, .(aice = mean(aice)), by = .(dataset, time = as.Date(round_date(time, "month")))] |>
  _[, aice := Anomaly(aice, year(time) %between% c(1981, 2011), na.rm = TRUE),
    by = .(month(time), dataset)] |>
  _[]

@@

hindast_extent_monthly <- hindcast_extent |> 
  copy() |> 
  _[, aice := aice - mean(aice[year(time) %between% c(1981, 2011)]),
    by = .(lag, model, time = update(time, year = 2000))] |>
  _[, .(forecast = mean(aice)),
    by = .(model, forecast_time, time = as.Date(round_date(time, "month")))] |> 
  _[, lag := months_difference(time, forecast_time)]

hindast_extent_monthly |> 
  _[, merge(.SD, monthly_extent, all = TRUE), by = .(model, lag)] |> 
  _[lag %in% (c(1, 3, 5, 7) - 1)] |>
  _[time %between% s2_range] |> 
  ggplot(aes(time)) +
  geom_line(aes(y = aice, group = dataset, color = dataset)) +
  geom_line(aes(y = forecast, color = model))  +
  geom_text(data = \(x) x[time %between% s1_range, 
                          rmse(aice, forecast, signif = 2, scale = 1e12),
                          by = .(lag, model)],
            aes(label = text),
            x = -Inf, y = Inf,
            vjust = 1.2, hjust = 0) +

  geom_text(data = \(x) x[time %between% s1_range, 
                          correlate(aice, forecast),
                          by = .(lag, model)],
            aes(label = text),
            vjust = 0, hjust = 1.2,
            x = Inf, y = -Inf) +

  scale_y_continuous(NULL,
                     labels = labels_extent) +
  scale_x_date(NULL, expand = c(0, 0)) +
  facet_grid(lag ~ model, labeller = labeller(lag = \(x) paste0("Month: ", x))) +
  scale_color_models +
  scale_fill_models
```

@@

obs_sd <- monthly_extent[time %between% s1_range,
                                   sd_ci(aice, na.rm = TRUE),
                                   by = .(month(time), dataset)] |> 
  _[, .(dataset, month, obs_sd = estimate)]

hindcast_extent_sd <- hindast_extent_monthly[time %between% s1_range, 
                       sd_ci(forecast, na.rm = TRUE), 
                       by = .(model, lag, month(time))]

hindcast_extent_sd |> 
  _[obs_sd, on = .NATURAL] |> 
  ggplot(aes(lag, estimate / obs_sd)) +
  # geom_rect(data = obs_sd, inherit.aes = FALSE,
  #            aes(ymin = low, ymax = high,  fill = dataset, xmin = -Inf, xmax = Inf),
  #               alpha = 0.5) +
  # geom_hline(data = obs_sd, aes(yintercept = estimate, color = dataset)) +
  geom_ribbon(aes(ymin = low/obs_sd, ymax = high/obs_sd, fill = model), alpha = 0.5) +
  geom_line(aes(color = model)) +

  scale_y_continuous("Standardised sea ice extent standard deviation") +
  scale_x_continuous("Lead time", breaks = c(0:9), expand = c(0, 0), 
                     trans = scales::reverse_trans()) +
  scale_color_models +
  scale_fill_models +
  facet_wrap(~ month, labeller = labeller(month = labels_month))

@@

aice_forecast_example <- c(lapply(c("S1", "S2"), \(model) {
  month_clim <- formatC(month(forecast_date), width = 2, flag = "0")

  hindcast(forecast_date, model, 1) |> 
    cdo_ydaysub(here::here(glue::glue("data/derived/climatology/{model}/{month_clim}/em.nc"))) |>
    cdo_execute() |> 
    ReadNetCDF("aice", subset = list(time = sample_date)) |> 
    _[, model := model]
  }),
  list(CDR() |> 
       anomalies() |>
       ReadNetCDF("aice", subset = list(time = sample_date)) |> 
       _[, model := "cdr"])) |> 
  rbindlist()

thickness_example <- c(S2 = "/g/data/ux62/access-s2/hindcast/raw_model/ice/hi/daily/e01/di_hi_{format(forecast_date, '%Y%m%d')}_e01.nc",
  S1 = "/g/data/ub7/access-s1/hc/raw_model/unchecked/ice/ice/daily/e01/di_ice_{format(forecast_date, '%Y%m%d')}_e01.nc") |> 
  purrr::imap(\(f, model) {
    f |> 
      glue::glue() |> 
      cdo_selname("hi") |> 
      remap_cdr() |> 
      cdo_execute(cache = TRUE) |> 
      ReadNetCDF("hi", subset = list(time = sample_date))

  }) |> 
  rbindlist(idcol = "model")

@@

```{r fig-forecast-example}
#| fig.cap: +S1 and +S2 hindcasts for the 2 of May 2008. Top row shows sea-ice concentration anomalies forecasted by each system and the observations. Bottom row shows forecasted sea-ice thickness and the difference between +S1 and +S2.
#| fig.height: 7
#| fig.width: 5.5
b <- seq(-0.5, 0.5, by = 0.1)
b <- b[b!= 0]

aice_forecast_example |> 
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = aice, fill = ..level..),
                    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)) +
  scale_fill_divergent_discretised("Sea-ice concentration anomaly",
                                   low = trans_pink,
                                   high = trans_blue) +
  geom_antarctica_fill + 
  geomcoord_antarctica +
  facet_wrap(~model, labeller = labeller(model = labels_models))  +
  wide_legend  +

thickness_example |> 
  _[, model := factor(model, levels = c("S1", "S2", "Difference"))] |> 
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(aes(z = hi, fill = ..level..), 
                    breaks = c(AnchorBreaks(exclude = 0, binwidth = 0.25)(c(0, 2)), Inf)) +
  # as.discretised_scale(scale_fill_viridis_c)("Sea ice thickness", 
  #                                            guide = guide_colorsteps(order = 9)) + 
  scale_fill_divergent_discretised("Sea-ice thickness",
                                   high = scales::muted("#5ab4ac"),
                                   guide = guide_colorsteps(order = 9)) +
  ggnewscale::new_scale_fill() +

  geom_contour_fill(data = \(x) x |> 
                      dcast(time + xgrid + ygrid ~ model, value.var = "hi") |> 
                      _[, model := factor("Difference", levels = c("S1", "S2", "Difference"))], 
                    aes(z = S1 - S2, fill = ..level..), breaks = c(-Inf, b, Inf)) +

  scale_fill_divergent_discretised("Sea-ice thickness difference",
                                   guide = guide_colorsteps(order = 0)) +
  geom_antarctica_fill + 
  geomcoord_antarctica +
  wide_legend +
  facet_wrap(~ model) +
  plot_layout(ncol = 1)

@@

mean_thickness_access <- \(forecast_time, model) {
  concentration <- hindcast(forecast_time, model = model, members = "em", variable = "aice")
  thickness <- hindcast(forecast_time, model = model, members = "em", variable = "hi")

  if (length(thickness) == 0) return(NA_character_)
  if (length(concentration) == 0) return(NA_character_)

@@

get_dir_part <- function(path, n = 1) {
  if (n == 1) return(basename(path))

@@

file <- "sithick_SImon_MCM-UA-1-0_historical_r1i1p1f1_gn_185001-201412.nc"
dates <- function(path) {
  file <- get_dir_part(path, 1)
  l <- 6+1+6+3
  n <- nchar(file)
  s <- n-l+1

  date_start <- substr(file, start = s, s + 5)
  date_end <- substr(file, start = s+7, n-3)
  list(start = lubridate::ym(date_start),
       end = lubridate::ym(date_end))
}

@@

```{r}
hindcast_thickness <- CJ(model = c("S2", "S1")) |>
  _[, .(forecast_time = get_forecast_times(model)), by = model] |> 
  _[mday(forecast_time) == 1] |> 
  _[, thickness := unlist(mirai_map(data.frame(forecast_time, model), 
                                    mean_thickness_access,
                                    hindcast = hindcast, mean_thickness = mean_thickness)[.progress])] |> 
  na.omit()

@@

giomas <- mean_thickness(giomas_concentration, giomas_thickness, giomas_cellarea) |> 
  cdo_ymonmean() |> 
  cdo_execute() |> 
  ReadNetCDF(c(thickness = "heff")) |> 
  _[, month := 1:12]

```
```{r}
hindcast_mean_thickness <- hindcast_thickness |> 
  _[, thickness |> cdo_monmean() |> cdo_execute(), by = .(model, forecast_time)] |>
  _[, metR::ReadNetCDF(V1, c(thickness = "hi")), by = .(model, forecast_time)] |> 
  _[, time := as.Date(update(time, day = 1))] |> 
  _[, lag :=  months_difference(time, forecast_time)] 
```

```{r}
cmip |> 
  _[model %in% unique(model)[1:10]] |> 
  _[, mean_thickness(cdo_sellonlatbox(concentration, 0, 360, -90, -40),
                     cdo_sellonlatbox(thickness, 0, 360, -90, -40)), 
            by = .(model, start, end)] |> 
    _[, .(thickness = V1 |> cdo_mergetime() |> cdo_execute()), by = .(model)] |> 
    _[, ReadNetCDF(thickness, "sithick")[, .(time, sithick)], by = .(model)] |> 
    _[, .(sithick = mean(sithick)), by = .(model, month(time))] |> 
    ggplot(aes(month, sithick)) +
    geom_line(aes(group = model, color = "CMIP6 models")) +
    geom_line(data = giomas, aes(y = thickness, color = "GIOMAS"), linewidth = 1.4) +
    scale_color_manual(values = c("CMIP6 models" = "black", GIOMAS =  "#1a7e21")) +
    scale_y_continuous("Mean sea-ice thickness") +
    scale_x_continuous(NULL, breaks = 1:12, labels = labels_month)
```

```{r}
hindcast_mean_thickness[time %between% s1_range,
                          average(thickness), 
                          keyby = .(lag, model, month(time))] |> 
  ggplot(aes(lag, estimate)) +
  geom_ribbon(aes(ymin = low, ymax = high, fill = model), alpha = 0.5) +
  geom_line(aes(color = model)) +
  # geom_smooth(method = "lm", aes(color = model)) +
  scale_color_models +
  scale_fill_models +
  scale_y_continuous("Mean sea-ice thickness") +
  scale_x_continuous("Lead time", breaks = c(0:9), expand = c(0, 0), 
                     trans = scales::reverse_trans()) +
  facet_wrap(~ month, labeller = labeller(month = labels_month))

@@

```{r clim_std}
clim_std <- datasets |> 
  _[1] |> 
  lapply(\(x) x |> 
           anomalies(climatology(x)) |> 
           cdo_sqr() |> 
           cdo_fldmean() |> 
           cdo_sqrt() |> 
           cdo_execute() |> 
           ReadNetCDF(c(climatology = "aice")) |> 
           _[, let(lat = NULL, lon = NULL)]
  ) |> 
  rbindlist(idcol = "obs_dataset") |> 
  _[, time := as.Date(time)] |> 
  _[climatology == 0, climatology := NA] |> 
  _[climatology < 0.05, climatology := NA]

@@

```{r t_test}
t_test <- function(x, y, ...) {

  test <- t.test(x, y, ...)

  list(low = test[["conf.int"]][1],
       high = test[["conf.int"]][2], 
       estimate = -diff(test$estimate), 
       p.value = test[["p.value"]]
  )
}

>>>>

```{r mirai_daemons}
if (!mirai::daemons_set()) {
  invisible(daemons(6))
}
```

@@

```{r range_max_lag}
range_max_lag <- hindcast_extent[
  mday(time) == 2,
  .SD[which.max(lag)],
  by = .(forecast_time, model)
] |>
  _[, .(max = max(lag), min = min(lag)), by = model]
```

@@

hindcast_extent[, .(model, forecast_time, member, time, Forecast = aice)] |>
  extent_daily[i = _, on = c("time")] |>
  _[, Difference := Forecast - aice] |>
  _[, aice := NULL] |>
  tidyfast::dt_pivot_longer(cols = c(Forecast, Difference)) |>
  _[,
    average(value),
    by = .(
      dataset,
      model,
      measure = name,
      forecast_time = update(forecast_time, year = 2001),
      lag = as.numeric(as.Date(time) - forecast_time)
    )
  ] |>
  _[, time := forecast_time + lag] |>
  _[, time2 := update(time, year = 2001)] |>
  _[, month := lubridate::month(forecast_time, label = TRUE)] |>
  _[, model := factor(model, levels = c("S1", "S2"))] |>
  _[, measure := factor(measure, levels = c("Forecast", "Difference"))] |>

  ggplot(aes(time2, estimate)) +
  geom_line(
    data = extent_climatology[dataset == which_dataset] |>
      _[, measure := factor("Forecast", levels = c("Forecast", "Difference"))],
    aes(x = as.Date(time), colour = dataset, group = dataset)
  ) +

  geom_hline(
    data = data.table(
      y = 0,
      measure = factor("Difference", levels = c("Forecast", "Difference"))
    ),
    aes(yintercept = 0)
  ) +
  geom_line(aes(
    group = interaction(forecast_time, time != time2),
    colour = model
  )) +
  geom_point(
    data = \(x) {
      x[
        mday(time) == 2,
        .SD[which.max(lag)],
        by = .(forecast_time, model, measure)
      ]
    },
    aes(fill = model),
    shape = 24,
    size = 2.2
  ) +
  geom_point(data = \(x) x[lag == 1], aes(fill = model), shape = 21, size = 2) +

  scale_x_date(
    NULL,
    date_breaks = "1 month",
    date_labels = "%b",
    expand = c(0, 0)
  ) +
  scale_y_continuous(NULL, labels = labels_extent) +
  scale_color_models +
  scale_fill_models +
  guides(colour = "none", fill = "none") +
  # scale_color_manual(values = cetcolor::cet_pal(12, name = "c2"),
  #                    aesthetics = c("fill", "color"), guide = "none") +
  facet_grid(
    measure ~ model,
    scales = "free_y",
    labeller = labeller(model = labels_models)
  ) +
  tag_facets(tag = "rc") +
  coord_cartesian(clip = "off")
```

@@

```{r extent-delta-compute}
n <- 366 * 3
w <- 11

dif_hindcast <- hindcast_extent |>
  copy() |>
  _[order(time)] |>
  _[, .(aice = mean(aice)), by = .(model, forecast_time, time)] |>
  _[,
    dif := c(NA, diff(aice) / diff(as.numeric(as.Date(time)))),
    by = .(model, forecast_time)
  ] |>
  _[,
    dif := frollmean(dif, n = w, align = "center"),
    by = .(model, forecast_time)
  ] |>
  na.omit() |>
  _[,
    average(dif),
    by = .(model, month(forecast_time), time = update(time, year = 2000))
  ]

dif_extent <- extent_daily |>
  _[order(time)] |>
  _[,
    dif := c(NA, diff(aice) / diff(as.numeric(as.Date(time)))),
    by = .(dataset)
  ] |>
  na.omit() |>
  _[, dif := frollmean(dif, n = w, align = "center"), by = .(dataset)] |>
  _[, average(dif), by = .(dataset, time = update(time, year = 2000))] |>
  _[dataset == which_dataset]

@@

dif_extent |>
  copy() |>
  ggplot(aes(time, estimate)) +
  geom_hline(yintercept = 0, color = "gray50", linewidth = 0.2) +
  geom_line(aes(color = dataset)) +
  geom_line(
    data = dif_hindcast,
    aes(color = model, group = interaction(model, month, month > month(time)))
  ) +
  scale_x_datetime(
    NULL,
    date_breaks = "1 month",
    date_labels = "%b",
    expand = c(0, 0)
  ) +
  scale_y_continuous(NULL, labels = \(x) {
    labels_extent(x, units = "M km²/day")
  }) +
  scale_color_models +
  coord_cartesian(xlim = as.POSIXct(c("2000-01-01", "2000-12-31"), tz = "UTC"))
```

```{r hindcast_clim}
zeropad <- function(x, n = 2) formatC(x, width = 2, flag = "0")
monnb <- function(d) {
  lt <- as.POSIXlt(as.Date(d, origin = "1900-01-01"))
  lt$year * 12 + lt$mon
}
mondf <- function(d1, d2) {
  monnb(d2) - monnb(d1)
}

@@

hindcast_clim <- lapply(setNames(months, months), \(month) {
  lapply(c("S1", "S2"), \(model) {
    here::here(glue::glue(
      "data/derived/climatology/{model}/{zeropad(month)}/em.nc"
    )) |>
      cdo_selmonth(month) |>
      cdo_execute() |>
      ReadNetCDF("aice") |>
      _[, model := model]
  }) |>
    rbindlist()
}) |>
  rbindlist(idcol = "forecast_month") |>
  _[, forecast_month := as.numeric(forecast_month)]

hindcast_clim <- hindcast_clim |>
  copy() |>
  _[,
    .(aice = mean(aice)),
    by = .(month(time), ygrid, xgrid, forecast_month, model)
  ] |>
  _[forecast_month == month]

@@

s <- lapply(c("S1", "S2"), \(x) {
  g <- hindcast_clim |>
    _[model == x] |>
    # dcast(time + xgrid + ygrid ~ model, value.var = "aice") |>
    # setnames(which_dataset, "obs") |>
    ggplot(aes(xgrid, ygrid)) +
    geom_contour_fill(
      aes(z = aice - obs, fill = after_stat(level)),
      breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)
    ) +
    scale_fill_divergent_discretised(
      paste0("Bias with respect to ", labels_models[which_dataset]),
      low = trans_pink,
      high = trans_blue
    ) +
    # geom_polygon(data = ice_shelves_nsicd, aes(X, Y, group = group),
    #              colour = "black", fill = "#FAFAFA") +
    geomcoord_antarctica +
    geom_antarctica_fill +
    facet_wrap(~ month + model, labeller = labeller(month = labels_month)) +
    wide_legend +
    tag_facets()

  print(g)
})

@@

monthly_extent <- extent_daily |>
  _[dataset == which_dataset] |>
  _[aice == 0, aice := NA] |>
  _[,
    .(aice = mean(aice)),
    by = .(dataset, time = as.Date(round_date(time, "month")))
  ] |>
  _[,
    aice := Anomaly(aice, year(time) %between% c(1981, 2011), na.rm = TRUE),
    by = .(month(time), dataset)
  ] |>
  _[]

@@

hindast_extent_monthly <- hindcast_extent |>
  copy() |>
  _[,
    aice := aice - mean(aice[year(time) %between% c(1981, 2011)]),
    by = .(lag, model, time = update(time, year = 2000))
  ] |>
  _[,
    .(forecast = mean(aice)),
    by = .(model, forecast_time, time = as.Date(round_date(time, "month")))
  ] |>
  _[, lag := months_difference(time, forecast_time)]

hindast_extent_monthly |>
  _[, merge(.SD, monthly_extent, all = TRUE), by = .(model, lag)] |>
  _[lag %in% (c(1, 3, 5, 7) - 1)] |>
  _[time %between% s2_range] |>
  ggplot(aes(time)) +
  geom_line(aes(y = aice, group = dataset, color = dataset)) +
  geom_line(aes(y = forecast, color = model)) +
  geom_text(
    data = \(x) {
      x[
        time %between% s1_range,
        rmse(aice, forecast, signif = 2, scale = 1e12),
        by = .(lag, model)
      ]
    },
    aes(label = text),
    x = -Inf,
    y = Inf,
    vjust = 1.2,
    hjust = 0
  ) +

  geom_text(
    data = \(x) {
      x[time %between% s1_range, correlate(aice, forecast), by = .(lag, model)]
    },
    aes(label = text),
    vjust = 0,
    hjust = 1.2,
    x = Inf,
    y = -Inf
  ) +

  scale_y_continuous(NULL, labels = labels_extent) +
  scale_x_date(NULL, expand = c(0, 0)) +
  facet_grid(
    lag ~ model,
    labeller = labeller(lag = \(x) paste0("Month: ", x))
  ) +
  scale_color_models +
  scale_fill_models
```

@@

obs_sd <- monthly_extent[
  time %between% s1_range,
  sd_ci(aice, na.rm = TRUE),
  by = .(month(time), dataset)
] |>
  _[, .(dataset, month, obs_sd = estimate)]

hindcast_extent_sd <- hindast_extent_monthly[
  time %between% s1_range,
  sd_ci(forecast, na.rm = TRUE),
  by = .(model, lag, month(time))
]

hindcast_extent_sd |>
  _[obs_sd, on = .NATURAL] |>
  ggplot(aes(lag, estimate / obs_sd)) +
  # geom_rect(data = obs_sd, inherit.aes = FALSE,
  #            aes(ymin = low, ymax = high,  fill = dataset, xmin = -Inf, xmax = Inf),
  #               alpha = 0.5) +
  # geom_hline(data = obs_sd, aes(yintercept = estimate, color = dataset)) +
  geom_ribbon(
    aes(ymin = low / obs_sd, ymax = high / obs_sd, fill = model),
    alpha = 0.5
  ) +
  geom_line(aes(color = model)) +

  scale_y_continuous("Standardised sea ice extent standard deviation") +
  scale_x_continuous(
    "Lead time",
    breaks = c(0:9),
    expand = c(0, 0),
    trans = scales::reverse_trans()
  ) +
  scale_color_models +
  scale_fill_models +
  facet_wrap(~month, labeller = labeller(month = labels_month))

@@

aice_forecast_example <- c(
  lapply(c("S1", "S2"), \(model) {
    month_clim <- formatC(month(forecast_date), width = 2, flag = "0")

    hindcast(forecast_date, model, 1) |>
      cdo_ydaysub(here::here(glue::glue(
        "data/derived/climatology/{model}/{month_clim}/em.nc"
      ))) |>
      cdo_execute() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := model]
  }),
  list(
    CDR() |>
      anomalies() |>
      ReadNetCDF("aice", subset = list(time = sample_date)) |>
      _[, model := "cdr"]
  )
) |>
  rbindlist()

thickness_example <- c(
  S2 = "/g/data/ux62/access-s2/hindcast/raw_model/ice/hi/daily/e01/di_hi_{format(forecast_date, '%Y%m%d')}_e01.nc",
  S1 = "/g/data/ub7/access-s1/hc/raw_model/unchecked/ice/ice/daily/e01/di_ice_{format(forecast_date, '%Y%m%d')}_e01.nc"
) |>
  purrr::imap(\(f, model) {
    f |>
      glue::glue() |>
      cdo_selname("hi") |>
      remap_cdr() |>
      cdo_execute(cache = TRUE) |>
      ReadNetCDF("hi", subset = list(time = sample_date))
  }) |>
  rbindlist(idcol = "model")

@@

```{r fig-forecast-example}
#| fig.cap: +S1 and +S2 hindcasts for the 2 of May 2008. Top row shows sea-ice concentration anomalies forecasted by each system and the observations. Bottom row shows forecasted sea-ice thickness and the difference between +S1 and +S2.
#| fig.height: 7
#| fig.width: 5.5
b <- seq(-0.5, 0.5, by = 0.1)
b <- b[b != 0]

aice_forecast_example |>
  ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(
    aes(z = aice, fill = ..level..),
    breaks = AnchorBreaks(binwidth = 0.1, exclude = 0)
  ) +
  scale_fill_divergent_discretised(
    "Sea-ice concentration anomaly",
    low = trans_pink,
    high = trans_blue
  ) +
  geom_antarctica_fill +
  geomcoord_antarctica +
  facet_wrap(~model, labeller = labeller(model = labels_models)) +
  wide_legend +

  thickness_example |>
    _[, model := factor(model, levels = c("S1", "S2", "Difference"))] |>
    ggplot(aes(xgrid, ygrid)) +
  geom_contour_fill(
    aes(z = hi, fill = ..level..),
    breaks = c(AnchorBreaks(exclude = 0, binwidth = 0.25)(c(0, 2)), Inf)
  ) +
  # as.discretised_scale(scale_fill_viridis_c)("Sea ice thickness",
  #                                            guide = guide_colorsteps(order = 9)) +
  scale_fill_divergent_discretised(
    "Sea-ice thickness",
    high = scales::muted("#5ab4ac"),
    guide = guide_colorsteps(order = 9)
  ) +
  ggnewscale::new_scale_fill() +

  geom_contour_fill(
    data = \(x) {
      x |>
        dcast(time + xgrid + ygrid ~ model, value.var = "hi") |>
        _[, model := factor("Difference", levels = c("S1", "S2", "Difference"))]
    },
    aes(z = S1 - S2, fill = ..level..),
    breaks = c(-Inf, b, Inf)
  ) +

  scale_fill_divergent_discretised(
    "Sea-ice thickness difference",
    guide = guide_colorsteps(order = 0)
  ) +
  geom_antarctica_fill +
  geomcoord_antarctica +
  wide_legend +
  facet_wrap(~model) +
  plot_layout(ncol = 1)

@@

mean_thickness_access <- \(forecast_time, model) {
  concentration <- hindcast(
    forecast_time,
    model = model,
    members = "em",
    variable = "aice"
  )
  thickness <- hindcast(
    forecast_time,
    model = model,
    members = "em",
    variable = "hi"
  )

  if (length(thickness) == 0) {
    return(NA_character_)
  }
  if (length(concentration) == 0) {
    return(NA_character_)
  }

@@

get_dir_part <- function(path, n = 1) {
  if (n == 1) {
    return(basename(path))
  }

@@

file <- "sithick_SImon_MCM-UA-1-0_historical_r1i1p1f1_gn_185001-201412.nc"
dates <- function(path) {
  file <- get_dir_part(path, 1)
  l <- 6 + 1 + 6 + 3
  n <- nchar(file)
  s <- n - l + 1

  date_start <- substr(file, start = s, s + 5)
  date_end <- substr(file, start = s + 7, n - 3)
  list(start = lubridate::ym(date_start), end = lubridate::ym(date_end))
}

@@

```{r}
hindcast_thickness <- CJ(model = c("S2", "S1")) |>
  _[, .(forecast_time = get_forecast_times(model)), by = model] |>
  _[mday(forecast_time) == 1] |>
  _[,
    thickness := unlist(mirai_map(
      data.frame(forecast_time, model),
      mean_thickness_access,
      hindcast = hindcast,
      mean_thickness = mean_thickness
    )[.progress])
  ] |>
  na.omit()

@@

giomas <- mean_thickness(
  giomas_concentration,
  giomas_thickness,
  giomas_cellarea
) |>
  cdo_ymonmean() |>
  cdo_execute() |>
  ReadNetCDF(c(thickness = "heff")) |>
  _[, month := 1:12]

```{r}
hindcast_mean_thickness <- hindcast_thickness |>
  _[,
    thickness |> cdo_monmean() |> cdo_execute(),
    by = .(model, forecast_time)
  ] |>
  _[,
    metR::ReadNetCDF(V1, c(thickness = "hi")),
    by = .(model, forecast_time)
  ] |>
  _[, time := as.Date(update(time, day = 1))] |>
  _[, lag := months_difference(time, forecast_time)]
```

```{r}
cmip |>
  _[model %in% unique(model)[1:10]] |>
  _[,
    mean_thickness(
      cdo_sellonlatbox(concentration, 0, 360, -90, -40),
      cdo_sellonlatbox(thickness, 0, 360, -90, -40)
    ),
    by = .(model, start, end)
  ] |>
  _[, .(thickness = V1 |> cdo_mergetime() |> cdo_execute()), by = .(model)] |>
  _[, ReadNetCDF(thickness, "sithick")[, .(time, sithick)], by = .(model)] |>
  _[, .(sithick = mean(sithick)), by = .(model, month(time))] |>
  ggplot(aes(month, sithick)) +
  geom_line(aes(group = model, color = "CMIP6 models")) +
  geom_line(
    data = giomas,
    aes(y = thickness, color = "GIOMAS"),
    linewidth = 1.4
  ) +
  scale_color_manual(values = c("CMIP6 models" = "black", GIOMAS = "#1a7e21")) +
  scale_y_continuous("Mean sea-ice thickness") +
  scale_x_continuous(NULL, breaks = 1:12, labels = labels_month)
```

```{r}
hindcast_mean_thickness[
  time %between% s1_range,
  average(thickness),
  keyby = .(lag, model, month(time))
] |>
  ggplot(aes(lag, estimate)) +
  geom_ribbon(aes(ymin = low, ymax = high, fill = model), alpha = 0.5) +
  geom_line(aes(color = model)) +
  # geom_smooth(method = "lm", aes(color = model)) +
  scale_color_models +
  scale_fill_models +
  scale_y_continuous("Mean sea-ice thickness") +
  scale_x_continuous(
    "Lead time",
    breaks = c(0:9),
    expand = c(0, 0),
    trans = scales::reverse_trans()
  ) +
  facet_wrap(~month, labeller = labeller(month = labels_month))

@@

```{r clim_std}
clim_std <- datasets |>
  _[1] |>
  lapply(\(x) {
    x |>
      anomalies(climatology(x)) |>
      cdo_sqr() |>
      cdo_fldmean() |>
      cdo_sqrt() |>
      cdo_execute() |>
      ReadNetCDF(c(climatology = "aice")) |>
      _[, let(lat = NULL, lon = NULL)]
  }) |>
  rbindlist(idcol = "obs_dataset") |>
  _[, time := as.Date(time)] |>
  _[climatology == 0, climatology := NA] |>
  _[climatology < 0.05, climatology := NA]

@@

```{r t_test}
t_test <- function(x, y, ...) {
  test <- t.test(x, y, ...)

  list(
    low = test[["conf.int"]][1],
    high = test[["conf.int"]][2],
    estimate = -diff(test$estimate),
    p.value = test[["p.value"]]
  )
}
